---
title: "PW6"
author: Mario GONZÁLEZ
date: "30/11/2022"
output:
    html_document:
        toc: yes
---

# Load the dataset
In this section we will load the _Spam_ dataset from the _kernlab_ package and analyze it.

```{r}
# Import the SPAM dataset and explore it.
library(kernlab)
data(spam)
summary(spam)
head(spam)
```

Afterwards we will split it into the training and test sets using a randomized seed to mantain
the random values.

```{r}
# Split the dataset into training and test sets using a random seed.
set.seed(13)
spam_samples <- caret::createDataPartition(
    spam$type,
    p = 0.7,
    list = FALSE
)
spam_train <- spam[spam_samples, ]
spam_train_scaled <- data.frame(
    scale(spam_train[, 1:57]),
    type = ifelse(spam_train$type == "spam", 1, 0)
)
spam_test <- spam[-spam_samples, ]
spam_test_scaled <- data.frame(
    scale(spam_test[, 1:57]),
    type = ifelse(spam_test$type == "spam", 1, 0)
)
```

# Train the models for classification

Now that the data is split and scaled, we can start training the models in order to compare their
performance.

## But before...

Before start training the models we will first proceed to create the results storage and create
a function to measure the categorical accuracy between the model predictions and the ground truth.

```{r}
# Create a function to return the binomial accuracy from two vectors.
binomial_accuracy <- function(real, prediction) {
    conf <- table(real, prediction)
    return(
        (conf[1, 1] + conf[2, 2]) /
        (conf[1, 1] + conf[1, 2] + conf[2, 1] + conf[2, 2])
    )
}
# Create the results dataframe.
spam_results <- data.frame(real_value = spam_test$type)
```

## Logistic regression

Now we will start training the logistic regression model and compare its results.

```{r}
spam_log <- glm(
    type ~ .,
    data = spam_train_scaled,
    family = "binomial"
)
summary(spam_log)
spam_results$logistic <- ifelse(
    test = predict(
        object = spam_log,
        newdata = spam_test_scaled,
        type = "response"
    ) >= 0.5,
    yes = 1,
    no = 0
)
binomial_accuracy(spam_results$real_value, spam_results$logistic)
plot(
    logistic ~ real_value,
    data = spam_results,
    main = "Logistic regression"
)
```

## Classification Tree

Now we will fit a classification tree using the training data and compare its results.

```{r}
# Fit a classification tree using the training set.
spam_tree <- rpart::rpart(type ~ ., spam_train_scaled)
summary(spam_tree)
rpart.plot::rpart.plot(spam_tree)
spam_results$tree <- ifelse(
    test = predict(
        object = spam_tree,
        newdata = spam_test_scaled
    ) >= 0.5,
    yes = 1,
    no = 0
)
binomial_accuracy(spam_results$real_value, spam_results$tree)
plot(
    tree ~ real_value,
    data = spam_results,
    main = "Classification tree"
)
```

## Bagging tree

Now we will fit a bagged tree and compare its results against the others.

```{r}
# Fit a Bagging model using the training set.
spam_bag <- ipred::bagging(type ~ ., spam_train_scaled)
summary(spam_bag)
spam_results$bag <- ifelse(
    test = predict(
        object = spam_bag,
        newdata = spam_test_scaled
    ) >= 0.5,
    yes = 1,
    no = 0
)
binomial_accuracy(spam_results$real_value, spam_results$bag)
plot(
    bag ~ real_value,
    data = spam_results,
    main = "Bagged tree"
)
```

## Random forest

Now we fit a random forest model and evaluate its results with the training data.

```{r}
# Fit a random forest model using the training set.
spam_forest <- randomForest::randomForest(type ~ ., spam_train_scaled)
summary(spam_forest)
spam_results$forest <- ifelse(
    test = predict(
        object = spam_forest,
        newdata = spam_test_scaled
    ) >= 0.5,
    yes = 1,
    no = 0
)
binomial_accuracy(spam_results$real_value, spam_results$forest)
plot(
    forest ~ real_value,
    data = spam_results,
    main = "Random forest"
)
```

## Gradient Boosting

Now we will train a Gradient Boosting Model for classification and evaluate its results
using the training data.

```{r}
spam_gbm <- gbm::gbm(type ~ ., data = spam_train_scaled)
summary(spam_gbm)
spam_results$gbm <- ifelse(
    test = predict(
        object = spam_gbm,
        newdata = spam_test_scaled
    ) >= 0.5,
    yes = 1,
    no = 0
)

binomial_accuracy(spam_results$real_value, spam_results$gbm)
plot(
    gbm ~ real_value,
    data = spam_results,
    main = "Gradient Boosting Model"
)
```

## Comparison of performances

In order to compare the performances of the models, we will plot the difference of the binomial
accuracy scores.

```{r}
spam_acc <- c(
    "logistic model" = binomial_accuracy(
        spam_results$real_value,
        spam_results$logistic
    ),
    "classification tree" = binomial_accuracy(
        spam_results$real_value,
        spam_results$tree
    ),
    "bagging tree" = binomial_accuracy(
        spam_results$real_value,
        spam_results$bag
    ),
    "random forest" = binomial_accuracy(
        spam_results$real_value,
        spam_results$forest
    ),
    "gbm" = binomial_accuracy(
        spam_results$real_value,
        spam_results$gbm
    )
)

barplot(
    spam_acc,
    main = "Binomial accuracy values for the classification models",
    las = 2,
    cex.names = 0.7
)
```

# Train and tune the models

In this section we will geenerate new training models and tune them using the _caret_ package
functions in order to fine-tune their parameters.

## But before...
We will generate new datasets containing factorized categorical values in order to represent
the target values.

```{r}
spam_ftrain_scaled <- data.frame(
    scale(spam_train[, 1:57]),
    type = spam_train$type
)
spam_ftest_scaled <- data.frame(
    scale(spam_test[, 1:57]),
    type = spam_test$type
)
```

## Logistic regression

Now we will train and fine-tune a logistic regression model using the _caret_ package for
fine-tuning.

```{r}
spam_glm_tuned <- caret::train(
    form = type ~ .,
    data = spam_ftrain_scaled,
    trControl = caret::trainControl(method = "cv", number = 5),
    method = "glm",
    family = "binomial"
)
spam_results$tuned_glm <- ifelse(
    predict(
        spam_glm_tuned,
        spam_ftest_scaled
    ) == "spam",
    1,
    0
)

binomial_accuracy(spam_results$real_value, spam_results$tuned_glm)
plot(
    tuned_glm ~ real_value,
    data = spam_results,
    main = "Tuned Logistic Regression"
)
```

## Random Forest
Now we will train and fine tune a Random Forest model using the training data.

```{r}
spam_rf_tuned <- caret::train(
    form = type ~ .,
    data = spam_ftrain_scaled,
    trControl = caret::trainControl(method = "oob"),
    method = "rf"
)
spam_results$tuned_rf <- ifelse(
    predict(
        spam_rf_tuned,
        spam_ftest_scaled
    ) == "spam",
    1,
    0
)

binomial_accuracy(spam_results$real_value, spam_results$tuned_rf)
plot(
    tuned_rf ~ real_value,
    data = spam_results,
    main = "Tuned Random Forest"
)
```

## Bagged Tree
Now we will train and fine-tune a Bagged Tree model using the training formatted data.

```{r}
spam_bag_tuned <- caret::train(
    form = type ~ .,
    data = spam_ftrain_scaled,
    trControl = caret::trainControl(method = "cv", number = 5),
    method = "treebag",
    bagControl = caret::bagControl(
        fit = spam_ftrain_scaled[1:57, ],
        predict = spam_ftrain_scaled$type,
        oob = TRUE
    )
)
spam_results$tuned_bag <- ifelse(
    predict(
        spam_bag_tuned,
        spam_ftest_scaled
    ) == "spam",
    1,
    0
)

binomial_accuracy(spam_results$real_value, spam_results$tuned_bag)
plot(
    tuned_bag ~ real_value,
    data = spam_results,
    main = "Tuned Bagged Tree"
)
```

## Gradient Boosting Model
Now we will train and fine tune a Gradient Boosting Model and evaluate its results using
the training data.

```{r}
spam_gbm_tuned <- caret::train(
    form = type ~ .,
    data = spam_ftrain_scaled,
    trControl = caret::trainControl(method = "cv", number = 5),
    method = "gbm"
)
spam_results$tuned_gbm <- ifelse(
    predict(
        spam_gbm_tuned,
        spam_ftest_scaled
    ) == "spam",
    1,
    0
)

binomial_accuracy(spam_results$real_value, spam_results$tuned_gbm)
plot(
    tuned_gbm ~ real_value,
    data = spam_results,
    main = "Tuned Gradient Boosting Model"
)
```

## Compare the results
Now we will compare the performance of each model using their binomial accuracy scores.

```{r}
spam_acc_tuned <- c(
    "logistic model" = binomial_accuracy(
        spam_results$real_value,
        spam_results$tuned_glm
    ),
    "bagging tree" = binomial_accuracy(
        spam_results$real_value,
        spam_results$tuned_bag
    ),
    "random forest" = binomial_accuracy(
        spam_results$real_value,
        spam_results$tuned_rf
    ),
    "gradient boosting" = binomial_accuracy(
        spam_results$real_value,
        spam_results$tuned_gbm
    )
)

barplot(
    spam_acc_tuned,
    main = "Binomial accuracy values for the tuned classification models",
    las = 2,
    cex.names = 0.7
)
```