---
title: "Machine Learning - PW5"
author: "Mario GONZÁLEZ GALINDO"
date: "28/11/2022"
output:
    html_document:
        toc: yes
---

In this practical work we will train and evaluate regression models using
decision trees.

# Load the dataset and shuffle it into training and test sets
Here we will load the **Boston** dataset from the _Mass_ library, which
will then be shuffled and splitted in half in order to form the training and
test sets that will be used for the rest of the project.
```{r}
library(MASS)
data(Boston)

boston_shuffled <- Boston[sample(1 : nrow(Boston)), ]
boston_test <- boston_shuffled[1 : (nrow(Boston) / 2), ]
boston_train <- boston_shuffled[(1 + nrow(Boston) / 2) : nrow(Boston), ]
```

# Fit different regression models and compare their performance
In this section we will train different regression models into the training set
and compare their performance with each other.

The models that will be evaluated for this data stand as following:

- [Regression tree](#regression-tree)
- [Linear Regression model](#linear-model)
- [Bagged tree](#bagged-tree)
- [Random Forest](#random-forest)
- [Gradient Boosting Model](#gradient-boosting-model)

## Before continuing...
In order to evaluate the models we will create two objects that we will use as
tools for the comparisoon of the models.
1. A function that evaluates the **RMSE** _(Root Mean Squared Error)_.
2. A data frame to store the prediction values of each model as well as the actual values.

```{r}
rmse <- function(vec_1, vec_2) sqrt(mean((vec_1 - vec_2)^2))
boston_results <- data.frame(real_value = boston_test$medv)
```

## Regression tree
In this section we will fit a regression tree model, which will be used to
predict the _medv_ variable from the dataset. In order to fit and train the
regression tree, we will use the _rpart_ package. Then we will make the
predictions for the test set using the model and visualize the _RMSE_.
```{r boston_tree, fig.show="hold", fig.width=4}
# Fit a regression tree to estimate the variable medv using the train dataset.
boston_tree <- rpart::rpart(medv ~ ., boston_train)
summary(boston_tree)
# Make predictions using the regression tree in the test dataset.
boston_results$tree <- predict(
    object = boston_tree,
    newdata = boston_test
)
rmse(boston_results$real_value, boston_results$tree)
plot(
    tree ~ real_value,
    data = boston_results,
    main = "Regression tree Predictions vs Real values",
    xlab = "Real values",
    ylab = "Predictions"
)
```

### Visualize the model
We can observe the complexity of the tree and the nodes needed to categorize and produce the prediction.
```{r}
plot(boston_tree)
rpart.plot::rpart.plot(boston_tree)
rpart::printcp(boston_tree)
rpart::plotcp(boston_tree)
```

## Linear model
Now we will fit a linear model as we know it and compare the results against
the regression tree.
```{r}
# Fit a linear model to compare its results against the regression tree.
boston_linear <- lm(medv ~ ., boston_train)
boston_results$linear <- predict(
    object = boston_linear,
    newdata = boston_test
)
rmse(boston_results$real_value, boston_results$linear)
plot(
    linear ~ real_value,
    data = boston_results,
    main = "Linear regression Predictions vs Real values",
    xlab = "Real values",
    ylab = "Predictions"
)
```

## Bagged tree
Now we will fit a Bagged tree using the _ipred_ package and evaluate its performance.
```{r}
#Fit a Bagged tree and compare its predictions against the others.
boston_bag <- ipred::bagging(medv ~ ., data = boston_train)
boston_results$bagging <- predict(
    object = boston_bag,
    newdata = boston_test
)
rmse(boston_results$real_value, boston_results$bagging)
plot(
    bagging ~ real_value,
    data = boston_results,
    main = "Bagging Predictions vs Real values",
    xlab = "Real values",
    ylab = "Predictions"
)
```

## Random Forest
Now we will fit a Random forest and evaluate its performance
```{r}
# Fit a Random Forest and compare its predictions against the other models.
boston_forest <- randomForest::randomForest(medv ~ ., data = boston_train)
boston_results$forest <- predict(
    object = boston_forest,
    newdata = boston_test
)
rmse(boston_results$real_value, boston_results$forest)
plot(
    forest ~ real_value,
    data = boston_results,
    main = "Random Forest Predictions vs Real values",
    xlab = "Real values",
    ylab = "Predictions"
)
```

### Predictors evaluation
Here we will obtain the importance of each predictor variable passed into the
Random Forest and order the predictors in order to retrieve the most important ones
used for the model.
```{r}
# Retrieve the Random Forest predictors and store them in a data frame
forest_predictors <- data.frame(
    randomForest::importance(boston_forest)
)
# Show the three most important predictors of the forest.
forest_predictors <- forest_predictors[
    order(-forest_predictors$IncNodePurity),
    , drop = FALSE
]
print(forest_predictors[1:3, , drop = FALSE])
randomForest::varImpPlot(boston_forest)
```

## Gradient Boosting Model
Now we will fit a Gradient Boosting Model and evaluate its performance.
```{r}
# Fit a Gradient Boosting model to the dataset and compare it
# against the other models.
boston_gbm <- gbm::gbm(medv ~ ., data = boston_train)
boston_results$gbm <- predict(boston_gbm, boston_test)
summary(boston_gbm)
rmse(boston_results$real_value, boston_results$gbm)
plot(
    gbm ~ real_value,
    data = boston_results,
    main = "Regression tree Predictions vs Real values",
    xlab = "Real values",
    ylab = "Predictions"
)
```

## Compare the performance of the models
```{r}
# Compare the RMSE of all the models fitted into the dataset.
boston_rmse <- c(
    "regression tree" = rmse(
        boston_results$real_value,
        boston_results$tree
    ),
    "linear model" = rmse(
        boston_results$real_value,
        boston_results$linear
    ),
    "bagging model" = rmse(
        boston_results$real_value,
        boston_results$bagging
    ),
    "random forest" = rmse(
        boston_results$real_value,
        boston_results$forest
    ),
    "gbm" = rmse(
        boston_results$real_value,
        boston_results$gbm
    )
)
barplot(
    boston_rmse,
    main = "RMSE values for the regression models",
    las = 2,
    cex.names = 0.7
)
```

```{r predictioin-plots, fig.width=4, fig.show="hold"}
plot(
    tree ~ real_value,
    data = boston_results,
    main = "Regression Tree model",
    xlab = "Real value",
    ylab = "Predictions"
)
plot(
    linear ~ real_value,
    data = boston_results,
    main = "Linear model",
    xlab = "Real value",
    ylab = "Predictions"
)
plot(
    bagging ~ real_value,
    data = boston_results,
    main = "Bagging model",
    xlab = "Real value",
    ylab = "Predictions"
)
plot(
    forest ~ real_value,
    data = boston_results,
    main = "Random Forest model",
    xlab = "Real value",
    ylab = "Predictions"
)
plot(
    gbm ~ real_value,
    data = boston_results,
    main = "GBM model",
    xlab = "Real value",
    ylab = "Predictions"
)
```